Reanalysis Step 2: Multiple Imputation

### 2025-08-30 reanalysis-2-multiple_imputation.qmd
implement fellows

@@

  # DR severity - use person_dr (created from No_DR/NPDR/PDR)
  "person_dr"  # Numeric 0/1/2
  # NOT dr_severity or person_dr_severity - derived from person_dr
)

@@

@@

@@

>>>>

## Create Fellow Eye Variables

```{r create-fellow-eye-vars}
# Create fellow eye variables by pairing eyes within patients
# This leverages correlation between eyes to improve imputation

# First, check data structure and create pairing
df_with_fellow <- df_for_imputation |>
  group_by(e_mrn_deidentified) |>
  mutate(
    n_eyes = n(),
    eye_index = row_number(),

    # Verify no patient has >2 eyes
    max_eyes_check = max(n_eyes)
  ) |>
  ungroup()

# Data check: Ensure no patient has >2 eyes
if (max(df_with_fellow$max_eyes_check) > 2) {
  stop("Found patients with >2 eyes. Check data quality.")
}
log_message(paste0("Maximum eyes per patient: ", max(df_with_fellow$max_eyes_check)))

# Create fellow variables
df_with_fellow <- df_with_fellow |>
  group_by(e_mrn_deidentified) |>
  mutate(
    # Fellow eye values (NA for single-eye patients)
    # Use [3 - eye_index] trick: eye 1 gets eye 2, eye 2 gets eye 1
    fellow_baseline_VA = if_else(n_eyes == 2, 
                                 baseline_VA_logMAR[3 - eye_index], 
                                 NA_real_),
    fellow_intermediate_VA = if_else(n_eyes == 2,
                                    intermediate_VA_logMAR[3 - eye_index],
                                    NA_real_),
    fellow_outcome_VA = if_else(n_eyes == 2,
                               outcome_VA_logMAR[3 - eye_index],
                               NA_real_)
    # Note: fellow_dr not needed since person_dr is already consistent within patients
  ) |>
  ungroup() |>
  select(-max_eyes_check, -eye_index)

# Patient distribution summary
patient_summary <- df_with_fellow |>
  group_by(e_mrn_deidentified) |>
  slice(1) |>  # One row per patient
  ungroup() |>
  summarise(
    total_patients = n(),
    single_eye_patients = sum(n_eyes == 1),
    bilateral_patients = sum(n_eyes == 2)
  )

log_message(paste0("Patient distribution - Total: ", patient_summary$total_patients,
                  ", Single eye: ", patient_summary$single_eye_patients,
                  ", Bilateral: ", patient_summary$bilateral_patients))

# Fellow data availability by cohort
fellow_check <- df_with_fellow |>
  filter(n_eyes == 2) |>
  group_by(cohort_id) |>
  summarise(
    n = n(),
    has_fellow_baseline = sum(!is.na(fellow_baseline_VA)),
    has_fellow_intermediate = sum(!is.na(fellow_intermediate_VA)),
    has_fellow_outcome = sum(!is.na(fellow_outcome_VA)),
    pct_fellow_baseline = round(100 * has_fellow_baseline / n, 1),
    pct_fellow_intermediate = round(100 * has_fellow_intermediate / n, 1),
    pct_fellow_outcome = round(100 * has_fellow_outcome / n, 1),
    .groups = 'drop'
  )

log_message("Fellow eye data availability by cohort:")
for(i in 1:nrow(fellow_check)) {
  log_message(paste0("  ", fellow_check$cohort_id[i], 
                    " - Baseline: ", fellow_check$pct_fellow_baseline[i], "%",
                    ", Intermediate: ", fellow_check$pct_fellow_intermediate[i], "%",
                    ", Outcome: ", fellow_check$pct_fellow_outcome[i], "%"))
}

# Eye correlation check (should be 0.5-0.9 typically)
correlation_check <- df_with_fellow |>
  filter(n_eyes == 2) |>
  summarise(
    cor_baseline = cor(baseline_VA_logMAR, fellow_baseline_VA, use = "complete.obs"),
    cor_intermediate = cor(intermediate_VA_logMAR, fellow_intermediate_VA, use = "complete.obs"),
    n_pairs_baseline = sum(!is.na(baseline_VA_logMAR) & !is.na(fellow_baseline_VA)),
    n_pairs_intermediate = sum(!is.na(intermediate_VA_logMAR) & !is.na(fellow_intermediate_VA))
  )

log_message(paste0("Fellow eye correlations - Baseline: ", round(correlation_check$cor_baseline, 3),
                  " (n=", correlation_check$n_pairs_baseline, " pairs)",
                  ", Intermediate: ", round(correlation_check$cor_intermediate, 3),
                  " (n=", correlation_check$n_pairs_intermediate, " pairs)"))

# Verification: Check fellow pairing worked correctly
pairing_check <- df_with_fellow |>
  filter(n_eyes == 2) |>
  group_by(e_mrn_deidentified) |>
  summarise(
    # Each eye should have the other eye's value as fellow
    check_baseline = baseline_VA_logMAR[1] == fellow_baseline_VA[2] & 
                    baseline_VA_logMAR[2] == fellow_baseline_VA[1],
    check_intermediate = (is.na(intermediate_VA_logMAR[1]) & is.na(fellow_intermediate_VA[2]) |
                         intermediate_VA_logMAR[1] == fellow_intermediate_VA[2]) &
                        (is.na(intermediate_VA_logMAR[2]) & is.na(fellow_intermediate_VA[1]) |
                         intermediate_VA_logMAR[2] == fellow_intermediate_VA[1]),
    .groups = 'drop'
  )

pairing_errors <- sum(!pairing_check$check_baseline | !pairing_check$check_intermediate, na.rm = TRUE)
if (pairing_errors > 0) {
  warning(paste0("Fellow eye pairing errors detected: ", pairing_errors, " patients"))
} else {
  log_message("Fellow eye pairing verification: PASSED")
}

# Use this enhanced dataset for imputation
df_for_imputation <- df_with_fellow
```

### Fellow Eye Data Summary

```{r fellow-summary-table}
# Display patient distribution
kable(patient_summary, caption = "Patient/Eye Distribution")

# Display fellow data availability
kable(fellow_check, 
      caption = "Fellow Eye Data Availability by Cohort (Bilateral Patients Only)",
      col.names = c("Cohort", "N Eyes", "Has Fellow Baseline", "Has Fellow Intermediate", 
                   "Has Fellow Outcome", "% Baseline", "% Intermediate", "% Outcome"))

# Display correlation summary
correlation_summary <- data.frame(
  Timepoint = c("Baseline", "Intermediate"),
  Correlation = c(correlation_check$cor_baseline, correlation_check$cor_intermediate),
  N_Pairs = c(correlation_check$n_pairs_baseline, correlation_check$n_pairs_intermediate)
)
kable(correlation_summary, 
      caption = "Inter-eye Correlations (Bilateral Patients)",
      digits = 3)
```

@@

  # DR severity - use person_dr (created from No_DR/NPDR/PDR)
  "person_dr",  # Numeric 0/1/2
  # NOT dr_severity or person_dr_severity - derived from person_dr

  # Fellow eye variables (for bilateral patients)
  "fellow_baseline_VA",
  "fellow_intermediate_VA",
  "fellow_outcome_VA",
  "n_eyes"  # Number of eyes per patient (1 or 2)
)

@@

# Fellow eye variables - don't impute (they're derived from existing data)
custom_methods["fellow_baseline_VA"] <- ""
custom_methods["fellow_intermediate_VA"] <- ""
custom_methods["fellow_outcome_VA"] <- ""
custom_methods["n_eyes"] <- ""  # Complete variable

@@

# FELLOW EYE PREDICTORS WITH TEMPORAL CONSTRAINTS
# Fellow baseline can predict all timepoints
if ("fellow_baseline_VA" %in% colnames(pred_matrix)) {
  pred_matrix["baseline_VA_logMAR", "fellow_baseline_VA"] <- 1
  pred_matrix["intermediate_VA_logMAR", "fellow_baseline_VA"] <- 1
  pred_matrix["outcome_VA_logMAR", "fellow_baseline_VA"] <- 1
}

# Fellow intermediate can predict intermediate and outcome (NOT baseline)
if ("fellow_intermediate_VA" %in% colnames(pred_matrix)) {
  pred_matrix["intermediate_VA_logMAR", "fellow_intermediate_VA"] <- 1
  pred_matrix["outcome_VA_logMAR", "fellow_intermediate_VA"] <- 1
  # CRITICAL: Prevent reverse-time prediction
  pred_matrix["baseline_VA_logMAR", "fellow_intermediate_VA"] <- 0
}

# Fellow outcome can predict outcome only (NOT earlier timepoints)
if ("fellow_outcome_VA" %in% colnames(pred_matrix)) {
  pred_matrix["outcome_VA_logMAR", "fellow_outcome_VA"] <- 1
  # CRITICAL: Prevent reverse-time prediction
  pred_matrix["baseline_VA_logMAR", "fellow_outcome_VA"] <- 0
  pred_matrix["intermediate_VA_logMAR", "fellow_outcome_VA"] <- 0
}

# Fellow variables are NEVER predicted (they're derived, not imputed)
fellow_vars <- c("fellow_baseline_VA", "fellow_intermediate_VA", "fellow_outcome_VA")
for (fvar in fellow_vars) {
  if (fvar %in% rownames(pred_matrix)) {
    pred_matrix[fvar, ] <- 0  # Not predicted by anything
  }
}

# n_eyes can be predictor but not predicted
if ("n_eyes" %in% rownames(pred_matrix)) {
  pred_matrix["n_eyes", ] <- 0  # Not predicted
  # n_eyes can help predict missingness patterns
  # Leave default (1) for using as predictor
}

# Log fellow predictor configuration
fellow_predictors_count <- c(
  baseline = sum(pred_matrix["baseline_VA_logMAR", grep("fellow", colnames(pred_matrix))]),
  intermediate = sum(pred_matrix["intermediate_VA_logMAR", grep("fellow", colnames(pred_matrix))]),
  outcome = sum(pred_matrix["outcome_VA_logMAR", grep("fellow", colnames(pred_matrix))])
)
log_message(paste0("Fellow predictors configured - Baseline: ", fellow_predictors_count["baseline"],
                  ", Intermediate: ", fellow_predictors_count["intermediate"],
                  ", Outcome: ", fellow_predictors_count["outcome"]))

@@

### Fellow Eye Correlation After Imputation

```{r fellow-correlation-check}
# Check that fellow eye correlations remain realistic after imputation
fellow_correlation_check <- lapply(1:n_imputations, function(i) {
  imp_data <- imputed_datasets[[i]]

  # Calculate correlations for bilateral patients
  cor_check <- imp_data |>
    filter(!is.na(n_eyes) & n_eyes == 2) |>
    summarise(
      cor_baseline = cor(baseline_VA_logMAR, fellow_baseline_VA, use = "complete.obs"),
      cor_intermediate = cor(intermediate_VA_logMAR, fellow_intermediate_VA, use = "complete.obs"),
      cor_outcome = cor(outcome_VA_logMAR, fellow_outcome_VA, use = "complete.obs"),
      n_baseline = sum(!is.na(baseline_VA_logMAR) & !is.na(fellow_baseline_VA)),
      n_intermediate = sum(!is.na(intermediate_VA_logMAR) & !is.na(fellow_intermediate_VA)),
      n_outcome = sum(!is.na(outcome_VA_logMAR) & !is.na(fellow_outcome_VA))
    )

  return(list(
    dataset = i,
    cor_baseline = round(cor_check$cor_baseline, 3),
    cor_intermediate = round(cor_check$cor_intermediate, 3),
    cor_outcome = round(cor_check$cor_outcome, 3),
    n_baseline = cor_check$n_baseline,
    n_intermediate = cor_check$n_intermediate,
    n_outcome = cor_check$n_outcome
  ))
})

# Display results
fellow_cor_df <- bind_rows(fellow_correlation_check)
log_message(paste0("Post-imputation fellow eye correlations - Outcome: ",
                  min(fellow_cor_df$cor_outcome, na.rm = TRUE), " to ",
                  max(fellow_cor_df$cor_outcome, na.rm = TRUE)))

# Check if correlations are in reasonable range (0.5-0.9 typically)
cor_warnings <- fellow_cor_df |>
  mutate(
    baseline_ok = cor_baseline >= 0.5 & cor_baseline <= 0.95,
    intermediate_ok = cor_intermediate >= 0.5 & cor_intermediate <= 0.95,
    outcome_ok = cor_outcome >= 0.5 & cor_outcome <= 0.95
  )

if (any(!cor_warnings$outcome_ok, na.rm = TRUE)) {
  warning("Some imputed datasets have fellow eye outcome correlations outside expected range (0.5-0.95)")
}
```

**Fellow eye correlations after imputation:**
```{r fellow-correlation-table}
kable(fellow_cor_df |> select(-starts_with("n_")), 
      col.names = c("Dataset", "Baseline", "Intermediate", "Outcome"),
      caption = "Inter-eye Correlations After Imputation (Bilateral Patients)",
      digits = 3)

# Also show the sample sizes
kable(fellow_cor_df |> select(dataset, starts_with("n_")), 
      col.names = c("Dataset", "N Baseline", "N Intermediate", "N Outcome"),
      caption = "Number of Paired Observations Used for Correlations")
```
